Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-12 paper2.qmd
section:6; process:write

<!-- Proposal: Make e_{I} and e_{J} fully orthogonal, a\
Decision: Although I can see the benefit of this, it would add several layers of explanations to the document or the code for that matter. I think this paper has already a lot of info. -->

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key issue in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference or hypothesis testing. The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct hypothesis tests. However, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
To address these issues, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. It begins by formulating a conceptual model represented by an SCM and a DAG. This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms. The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. In particular, it enables the analysis of CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when researchers seek to conduct statistical inference or hypothesis testing (refer to @sec-theory-issue2). These methodological innovations potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical conclusions.

@@

<!-- 2. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.

@@

<!-- Challenges -->

Despite its potential, this study presents several challenges that researchers must navigate. First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.

Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts. We have argued that the model's ability to estimate traits accurately and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.

Third, applying this approach requires a solid understanding of Causal inference. For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity. Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.

Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods. This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.

>>>>

<!-- Proposal: Make e_{I} and e_{J} fully orthogonal,
Decision: Although I can see the benefit of this, it would add several layers of explanations to the document or the code for that matter. I think this paper has already a lot of info. -->

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. It begins by formulating a conceptual model represented by an SCM and a DAG. This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms. The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. In particular, it enables the analysis of CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when researchers seek to conduct statistical inference or hypothesis testing (refer to @sec-theory-issue2). These methodological innovations potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical conclusions.

@@

<!-- 2. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two additional advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.

@@

<!-- Challenges -->
<!-- 1. Epistemiology  -->
Despite its potential, this study presents several challenges that researchers must navigate. First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.

<!-- 2. Research context is important  -->
Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts. We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.

<!-- 3. Causal inference knowledge -->
Third, applying this approach requires a solid understanding of Causal inference. For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity. Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.

<!-- 4. PPL and Bayesian Inference -->
Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods. This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand. While such translation is not exclusive to Bayesian methods--frequentist approaches can also be used--we argue that Bayesian methods offer a more accessible pathway,  as they reduce the need for additional expertise in optimization theory and techniques.
